---
title: "BSMM-lab-1"
subtitle: "BSMM 8740 Fall 2023"
author: "Maaz Saeed | GitHub: healyyyyyy"
date: "2023-09-22"
format: html
editor: visual
self-contained: true
---

## Setup

Load packages and data:

```{r load-pkg-data}
#| message: false
# install the package 'Lahman'
#install.packages('Lahman')
# load the package
library(Lahman)
library(dplyr)
library(magrittr)
```

## Exercises

### Exercise 1

```{r glimpse-data}
dplyr::glimpse(Teams)
```

The `Teams` dataset has 3,015 observations and 48 variables.

### Exercise 2
```{r Mets 2004-2012}
mets <- Teams  %>% 
  dplyr::filter(teamID == "NYN")
my_mets <- mets %>% 
  dplyr::filter(yearID > 2003 & yearID < 2013)
my_mets %>% 
  dplyr::select(teamID, yearID, W, L)
```
The win loss records for the New York Mets, for the years 2004 - 2012 are displayed above.

### Exercise 3
```{r gathering runs scored and allowed}
mets_ben <- Teams |>
  dplyr::select(teamID, yearID, G, W, L, R, RA,) |>
  dplyr::filter(teamID == "NYN" & yearID %in% 2004:2012)
#mets_ben

mets_ben <- mets_ben |>
  dplyr::rename(RS = R)    # new name = old name
mets_ben
```
The runs scored and allowed are displayed above for the years 2004 - 2012.

...

### Exercise 4
```{r adding wpct column}
mets_ben <- mets_ben |>
  dplyr::mutate(WPct = 1 / (1 + (RA/RS)^2))
mets_ben

mets_ben <- mets_ben |>
  dplyr::mutate(W_hat = WPct*G)
mets_ben
```
WPct = Win Percentage, W_hat = Expected number of wins.

...

### Exercise 5
```{r}
temp_var <- Teams |>
  dplyr::select(teamID, yearID, G, W,  R, RA) |>
  dplyr::filter(teamID == "NYN")

temp_var <- temp_var |>
  dplyr::rename(RS = R)


temp_var <- temp_var |>
  dplyr::mutate(WPct = 1 / (1 + (RA/RS)^2))


temp_var <- temp_var |>
  dplyr::mutate(W_hat = WPct*G)

temp_var <- temp_var |>
  dplyr::select(teamID, yearID, G, W,  RS, RA, WPct, W_hat) |>
  dplyr::filter(teamID =='NYN' & W >= as.integer(W_hat))
temp_var
```
They performed better in 30 seasons and worse in 31
(can be verified by total - better)

Or replacing the '>=' operator with '<' in the last line of this code chunk.

...

### Exercise 6
```{r}
dplyr::arrange(mets_ben, desc(W))
```
Above are the seasons ranked by the number of games won in descending order from top to bottom (highest to lowest).

...

### Exercise 7
```{r}
temp_again <- Teams |>
  dplyr::select(teamID, yearID, G, W,  R, RA) |>
  dplyr::filter(teamID == "NYN")

temp_again <- temp_again |>
  dplyr::rename(RS = R)

temp_again <- temp_again |>
  dplyr::mutate(WPct = 1 / (1 + (RA/RS)^2))

temp_again <- temp_again |>
  dplyr::mutate(W_hat = WPct*G)

temp_again <- temp_again |>
  dplyr::select(teamID, yearID, G, W, WPct, W_hat) |>
  dplyr::mutate(Diff = W - W_hat) |>
  dplyr::arrange(desc(Diff))
temp_again



```
The above are the seasons ranked in descending order by difference of expected wins and actual wins (top to down, best to worst).

### Exercise 7 Contd.
```{r}
mets_ben <-mets_ben %>% 
  dplyr::mutate(Diff = W -W_hat) |>
  dplyr::arrange(desc(Diff))
mets_ben
```
The above are the seasons ranked in descending order (top to down, best to worst), while Ben worked with the Mets.


### Exercise 8
```{r}
mets_ben |>
  dplyr::summarize(
    num_years = n(), 
    total_W = sum(W), 
    total_L = sum(L), 
    total_WPct = mean(WPct), 
    sum_resid = sum(Diff)
  )
```
In these nine years, the Mets had a combined record of 728 wins and 730 losses, for an overall winning percentage of approximately ~50.66%.


### Exercise 9
```{r}
#
#library(ggplot)
#ggplot(data = Ikea, aes(x = width, y = price)) +
#  geom_point() + 
#  labs(
#    x = "width", 
#    y = "price", 
#    title = "Ikea Furniture Width-Price"
#    )
```

### Exercise 10
```{r}
library(dplyr)
mets_ben <- mets_ben %>%
  dplyr::mutate(
    gm = case_when(
      yearID == 2004 ~ "Jim Duquette",
      yearID >= 2005 & yearID <= 2010 ~ "Omar Minaya",
      yearID >= 2011 ~ "Sandy Alderson"
    )
  )
mets_ben
```
Data divided into 3 groups (3 managers).

### Exercise 11
```{r manager grouping}
 
manager_summary <- mets_ben %>%
  group_by(gm) %>%
  summarize(
    num_years = n(),
    total_W = sum(W),
    total_L = sum(L),
    total_WPct = mean(WPct),
    sum_resid = sum(Diff)
  )

# Print the manager summary
manager_summary
```
Data grouped by manager name.
